---
title: "MMA Analytics"
output: md_document
#runtime: shiny
---

```{r include = FALSE}
setwd("C:/Users/Bashk/Desktop/Python/Projects/UFC")

library(Rmisc)
library(vcd)
library(viridis)
library(data.table)
library(ggplot2)
library(tidyr)
library(stringr)
library(stringi)
library(albersusa)
library(dplyr)
library(ggmap)
library(rworldmap)
library(knitr)


color_palette = 'RdBu'
sunshine_over_glacier <- c("#004056", "#2C858D", "#74CEB7", "#C9FFD5", "#FFFFCB")

fights <- read.csv("Data/Fights2.csv")
finishes <- read.csv("Data/Cat.csv")
fights <- merge(fights, finishes, by = "method_d", all = TRUE)

```
An analysis of 23 years of MMA outcomes in the Ultimate Fighting Championship (UFC)

There was a total of 3,569 fights between 1,561 fighters. The fights ranged from November 12<sup>th</sup>, 1993 up to February 21<sup>st</sup>, 2016. 

The complete R code for this project can be found on my [Github Site](www.github.com/bashkeel). The data was scraped by a 3rd party from the Sherdog website. It can be found [here](https://www.reddit.com/r/datasets/comments/47a7wh/ufc_fights_and_fighter_data/).

<br>

## Striking versus Grappling: How do fighters win in MMA?
Most fights are won through striking, followed by chokes and jointlocks. Regardless of the method of finish, a significant majority of fights are won within the first round.

If a fight doesn't end within the designated number of rounds (3 round for normal fights, 5 for championship or main event fights), the decision goes to the judges score card. Decision wins were excluded from this graph, but an analysis of decision wins can be found in the supplementary section. No Contest wins were also excluded. The resulting list of finishes included 2,044 fights.

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}

finishes_df <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, round) %>% 
  count(Category, round, sort = T) %>%
  filter(Category != "Decision") %>%
  mutate(Round = round)

finishes_catalogue <- finishes_df %>% group_by(Category) %>% summarise(total = sum(n)) %>% arrange(desc(-total))
finishes_df$Category <- factor(finishes_df$Category, levels=as.vector(finishes_catalogue$Category))

ggplot(finishes_df, aes(x=Category, y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Win Method", y="Count") + ggtitle("Method of Win") +
  scale_fill_manual(values = (sunshine_over_glacier)) + theme_minimal()

finishes_table <- dcast(finishes_df, Category~round, value.var = 'n', fill = 0)
finishes_table <- finishes_table[order(rowSums(finishes_table[,2:6]), decreasing = TRUE),]

kable(finishes_table, row.names = F, align='l')
```


<br>

## Fight Like a Girl: Do gender differences exist in how MMA fights are won?
In both men and women's MMA fights, an overwhelming majority of fights end in a decision. Faithful to the original analysis of finishes, strikes remain the dominant way of ending a fight, followed by chokes, then joint locks.

More men win fights with strikes (M: 33%, F: 25%) and choking (M: 16%, F: 14%), while women are more likely to end a fight with joint locks (F: 8%, M: 5%). Overall, women are more likely to continue fighting through all the rounds and end with a decision (F: 52%, M: 42%), rather than knocking out or submitting their opponents (M: 54%, F: 47%).

There were 3,445 men's MMA fight and 102 women's MMA fights in the dataset. Due to the large disparty between the two, the analysis of finishes required percentages rather than raw numbers. 

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}

finishes_by_gender <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, Gender) %>% 
  count(Category, Gender, sort = T) %>%
  group_by(Gender) %>% 
  mutate(Percentage = as.numeric(format(round(n/sum(n)*100,2), nsmall = 2))) %>%
  dcast(Gender~Category, value.var='Percentage', fill=0) %>%
  melt(id.vars = c('Gender'), variable.name = 'Category', value.name = 'Percentage') 

ggplot(finishes_by_gender, aes(x=reorder(Category, Percentage), y=Percentage, fill=Gender)) +
  geom_bar(position = "dodge", stat='identity', col='black', width=0.85) + coord_flip() +
  scale_fill_manual(values=c(sunshine_over_glacier[1], sunshine_over_glacier[4])) + 
  labs(x="Win Method", y="Percantage of Fights") + theme_minimal() +
  ggtitle("Differences in Finishes by Gender")

gendered_finish_catalogue <- finishes_by_gender %>%
  group_by(Category) %>% summarise(average = mean(Percentage)) %>%
  arrange(desc(average)) %>% dplyr::select(Category)

gendered_finishes_table <- dcast(finishes_by_gender, Gender~Category, value.var='Percentage', fill=0)
gendered_finishes_table <- gendered_finishes_table[,c("Gender", as.vector(gendered_finish_catalogue$Category))]
kable(gendered_finishes_table, row.names = F, align='l')

```

<br>

## Differences in decision wins by gender
Overall, men's fights end in unanimous decisions more often in unianimous decisions, whereby all three judges score the fight in favor of one fighter. On the other hand, women's fights end slightly more often in split or majority decisions, whereby two judges score the fight for the winning fighter while the third judge scores it for the losing fighter (split decision) or as a draw (majority decision).

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
decisions_by_gender <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, Gender) %>%
  filter(method_d %in% c("Unanimous", "Split", "Majority") & Category == "Decision") %>%
  count(method_d, Gender, sort = T) %>%
  group_by(Gender) %>% 
  mutate(Percentage = as.numeric(format(round(n/sum(n)*100,2), nsmall = 2))) %>%
  dcast(Gender~method_d, value.var='Percentage', fill=0) %>%
  melt(id.vars = c('Gender'), variable.name = 'Decision Method', value.name = 'Percentage')


ggplot(decisions_by_gender, aes(x=reorder(`Decision Method`, Percentage), y=Percentage, fill=Gender)) +
  geom_bar(position = "dodge", stat='identity', col='black', width=0.85) + coord_flip() +
  scale_fill_manual(values=c(sunshine_over_glacier[1], sunshine_over_glacier[4])) + 
  labs(x="Win Method", y="Percentage of Fights") + theme_minimal()

gendered_decision_table <- dcast(decisions_by_gender, Gender~`Decision Method`, value.var='Percentage', fill=0)
gendered_decision_table <- gendered_decision_table[,c("Gender", "Unanimous","Split", "Majority")]
kable(gendered_decision_table, align='l')

```

<br>

## Is Ring Rust Real?: 
Partially exaggerate, partially misunderstood, or partially true?

The graph below shows the average time fighters took between fights, and whether the fighter won his/her fight. Fighters who took longer between fights were statistically more likely to lose their next fight (p-value < 0.05). 

Despite the statistical significance, there is only a miniscule difference in average time between wins and losses (~3 weeks) and a relatively low variability (standard error) in both groups. Therefore, I would not conclude an additional three weeks of rest between fights will make or break a fight!

Fighters that fought multiple times in a single event were excluded. Statistical significance was determined using a two-sample Student's t-test. 

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
ringrust_df <- fights[,c("event_date", "f1fid", "f2fid")]
ringrust_wins <- setDT(ringrust_df[,1:2]) %>% mutate(Result = "Win") %>% rename(Fighter_ID = f1fid)
ringrust_losses <- setDT(ringrust_df[,c(1,3)]) %>% mutate(Result = 'Loss') %>% rename(Fighter_ID = f2fid)

ringrust_df <- rbind(ringrust_wins, ringrust_losses) %>% 
  mutate(event_date = as.Date(event_date, format = '%m/%d/%Y')) %>% 
  arrange(event_date, desc(event_date)) %>%
  group_by(Fighter_ID) %>%
  mutate(LastFight = event_date - lag(event_date, default = event_date[1])) %>%
  filter(LastFight != 0) %>%
  mutate(LastFight = as.numeric(LastFight))

ringrust_barplot <- summarySE(ringrust_df, measurevar = "LastFight", groupvars="Result")
ggplot(ringrust_barplot, aes(x=Result, y=LastFight, fill=Result)) +
  geom_bar(stat='Identity', col="black") +
  geom_errorbar(aes(ymin=LastFight, ymax=LastFight+se), width=0.2) +
  scale_fill_manual(values=c(sunshine_over_glacier[1], sunshine_over_glacier[4])) +
  ylab('Days Since Last Fight') + ggtitle("Ring Rust") + theme_minimal()

pval <- t.test(as.numeric(LastFight)~Result, data=ringrust_df)
pval <- data.frame(pval$estimate[[1]], pval$estimate[[2]], formatC(pval$p.value, format = "e", digits = 2))

colnames(pval) <- c("Loss: Mean Days", "Win: Mean Days", "P-value")

kable(pval, row.names = F, align='l')
```

<br>

## Is There a Hometown Advantage in MMA?
Whether it's psychological or physical, fights that take place in a fighter's home state are more likely to go in their favor. (Hometown: 57%, Away: 50%)

Hometowns were determined by the city that the fighter currently fights and trains out of, rather than their birthplace. 

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE, message=FALSE}

loc <- read.csv("Data/Areas2.csv")
fighters <- read.csv("Data/Fighters.csv")
fighters_loc <- fighters %>%
  mutate(Hometown = paste0(locality, ", ", country), Fighter_ID = fid) %>%
  dplyr::select(Hometown, Fighter_ID, Gender)


location_df <- setDT(fights) %>% dplyr::select(event_place,  f1fid, f2fid)
location_wins <- location_df[,c(1,2)] %>% mutate(Result = "Win") %>% dplyr::rename(Fighter_ID = f1fid)
location_losses <- location_df[,c(1,3)] %>% mutate(Result = "Loss") %>% dplyr::rename(Fighter_ID = f2fid)

location_df <- rbind(location_wins, location_losses) %>% setDT %>%
  mutate_if(is.factor, as.character) %>% rowwise() %>%
  mutate(Location = paste0(unlist(strsplit(event_place, split=", "))[2], ", ",
                           unlist(strsplit(event_place, split=", "))[3])) %>%
  dplyr::select(-event_place) %>%
  left_join(fighters_loc) %>% rowwise() %>%
  mutate(Hometown = paste0(unlist(strsplit(Hometown, split=", "))[1], ", ",
                           unlist(strsplit(Hometown, split=", "))[2])) %>%
  mutate(Local = as.character(Location == Hometown)) %>%
  mutate(Local=dplyr::recode(Local, `TRUE` = 'Home', `FALSE` = "Away")) %>%
  mutate(Result =  factor(Result, levels=c("Win", "Loss"))) %>%
  mutate(Local = factor(Local,levels=c("Home", 'Away')))


wins_by_hometown <- round(prop.table(table(location_df$Local,location_df$Result, dnn = list("Fighter", "Fight Result")),1)*100,1)

mosaic(wins_by_hometown, shade = TRUE, direction = "v", pop = FALSE, 
       gp = gpar(fill=matrix(c(sunshine_over_glacier[4], sunshine_over_glacier[1]), 2, 2)))
labeling_cells(text = as.table(wins_by_hometown), margin = 0)(as.table(wins_by_hometown))

kable(wins_by_hometown, align='l')
```

<br>

## The Land of Savages: Where do most UFC fighters come from?
A quick overview of fighters locality in the UFC. Due to the large disparity in numbers between fighters on the UFC roster from the US compared to other countries, the two groups were examined individually.

Within the US, the vast majority of fighters come from California (Eureka!). Outside of the US, the second largest market for UFC fighters is Brazil, followed by Canada, the UK, and Japan.

<br>
```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}

US_fighters <- setDT(fighters) %>%
  dplyr::select(fid, locality, country) %>%
  filter(country %in% c("United States", "USA")) %>%
  mutate_if(is.factor, as.character) %>% rowwise() %>%
  mutate(id = unlist(strsplit(locality, split=", "))[2]) 

US_fighters$id <- revalue(US_fighters$id,
                             c("Lousiana" = "Louisiana","Massachusets" = "Massachusetts",
                               "Massachusettes" = "Massachusetts", "Phildelphia" = "Pennsylvania"))

US_fighters_count <- US_fighters %>% count(id) %>% arrange(desc(n))

us <- usa_composite()
us_map <- fortify(us, region="name")

ggplot() + geom_map(data=us_map, map=us_map,
                    aes(x=long, y=lat, map_id=id),
                    color="#2b2b2b", size=0.1, fill='grey') + theme_void() +
  geom_map(data=US_fighters_count, map=us_map, aes(fill=n, map_id = id), color='black', size=0.15) +
  scale_fill_gradientn(colors=rev(sunshine_over_glacier), name='', breaks=c(1,60,120,185), labels=c(0,60,120,185)) + coord_map() +
  ggtitle("Number of Fighters Per US State") + theme_void()


US_table <- US_fighters_count[1:10,]
colnames(US_table) <- c("State", "Number of Fighters")
kable(US_table[1:10,], row.names = F, align = 'l', caption = "Top 10 States with most UFC Fighters")
```


<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE, message=FALSE}

#library(rgeos)
#library(raster)

origin <- fighters %>% count(country, sort = T)

ggmap <- invisible(joinCountryData2Map(origin, joinCode='NAME', nameJoinColumn = "country"))
countrymatch_failure <- setdiff(origin$country, as.vector(na.omit(ggmap$country)))

fighters$country <- revalue(fighters$country, 
                          c("England" = "United Kingdom", "USA" = "United States", "Holland" = "Netherlands",
                            "Northern Ireland" = "United Kingdom", "Scotland" = "United Kingdom", "Finnland" = 'Finland'))

#Recount & Remove US Fighters
NonUS <- fighters %>% 
  count(country, sort = T) %>% 
  filter(!country %in% c("United States", ""))

world_map<-map_data("world")
ggplot() + geom_map(data=world_map, map=world_map,
                    aes(x=long, y=lat, map_id=region),
                    color="#2b2b2b", size=0.1, fill='grey') + theme_void() +
  geom_map(data=NonUS, map=world_map, aes(fill=n, map_id = country), color='black', size=0.15) +
  scale_fill_gradientn(colors=rev(sunshine_over_glacier), name='', breaks=c(1,50,100,150,199), labels=c(1,50,100,150,200)) +
  ggtitle("Where do Non-US MMA Fighters Come From?") + 
  expand_limits(x = world_map$long, y = world_map$lat) + coord_equal()


NonUS_table <- NonUS[1:10,]
colnames(NonUS_table) <- c("Country", "Number of Fighters")
kable(NonUS_table, row.names = F, align = 'l', caption = "Top 10 Countries with most UFC Fighters (Excluding USA)")
```

<br>
<br>


# Supplementary Data

<br>

### Wins by Strikes

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
strikes <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "Strike") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round = round)

strikes_order <- aggregate(strikes$n, by=list(Strikes = strikes$method_d), FUN=sum) %>% 
  arrange(x) %>% filter(x > 3) %>% pull(Strikes)


ggplot(strikes, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Joint Lock Method", y="Count") + ggtitle("Types of Strike Finishes") +
  scale_fill_manual(values = sunshine_over_glacier) + theme_minimal() +
  scale_x_discrete(limits=strikes_order)


strikes_table <- aggregate(strikes$n, by=list(Strikes = strikes$method_d), FUN=sum) %>% 
  filter(x > 3) %>% arrange(desc(x)) %>%
  setNames(c("Strikes", "Count"))
kable(strikes_table, row.names = F, align='l')
```

<br>

### Wins by Choke

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
Choke <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "Choke") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round = round)

Choke_order <- aggregate(Choke$n, by=list(Choke = Choke$method_d), FUN=sum) %>% 
  arrange(x) %>% filter(x > 3) %>% pull(Choke)

ggplot(Choke, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Win Type", y="Count") + ggtitle("Types of Choke Finishes") +
  scale_fill_manual(values = sunshine_over_glacier) + theme_minimal() +
  scale_x_discrete(limits=Choke_order)

Choke_table <- aggregate(Choke$n, by=list(Choke = Choke$method_d), FUN=sum) %>% 
  filter(x > 3) %>% arrange(desc(x)) %>%
  setNames(c("Choke", "Count"))
kable(Choke_table, row.names = F, align='l')

```

<br>

### Wins by Joint Lock

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
joint_locks<- setDT(fights)%>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "JointLock") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round=round)

jointlock_order <- aggregate(joint_locks$n, by=list(Jointlock = joint_locks$method_d), FUN=sum) %>% 
  arrange(x) %>% filter(x > 3) %>% pull(Jointlock)

ggplot(joint_locks, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Joint Lock Method", y="Count") + ggtitle("Types of Joint Lock Finishes") +
  scale_fill_manual(values = sunshine_over_glacier) + theme_minimal() +
  scale_x_discrete(limits=jointlock_order)

joint_locks_table <- aggregate(joint_locks$n, by=list(joint_locks = joint_locks$method_d), FUN=sum) %>% 
  filter(x > 3) %>% arrange(desc(x)) %>%
  setNames(c("Joint Lock", "Count"))
kable(joint_locks_table, row.names = F, align='l')
```

<br>

### Wins by Stoppage

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
Stoppage <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "Stoppage") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round = round)

Stoppage_order <- aggregate(Stoppage$n, by=list(Stoppage = Stoppage$method_d), FUN=sum) %>% 
  arrange(x) %>% pull(Stoppage)

ggplot(Stoppage, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Win Type", y="Count") + ggtitle("Types of Stoppages") +
  scale_fill_manual(values = sunshine_over_glacier) + theme_minimal() +
  scale_x_discrete(limits=Stoppage_order)


Stoppage_table <- aggregate(Stoppage$n, by=list(Stoppage = Stoppage$method_d), FUN=sum) %>% 
  arrange(desc(x)) %>%
  setNames(c("Stoppage", "Count"))
kable(Stoppage_table, row.names = F, align='l')
```

<br>

### Wins by Miscellaneous Methods

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
misc <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "MISC") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round = round)

misc_order <- aggregate(misc$n, by=list(Misc = misc$method_d), FUN=sum) %>% 
  arrange(x) %>% pull(Misc)

ggplot(misc, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Win Type", y="Count") + ggtitle("Types of Miscellaneous Finishes") +
  scale_fill_manual(values = sunshine_over_glacier[c(1,3,4,5)]) + theme_minimal() +
  scale_x_discrete(limits=misc_order) + scale_y_continuous(breaks=c(0,5,10))

misc_table <- aggregate(misc$n, by=list(misc = misc$method_d), FUN=sum) %>% 
  arrange(desc(x)) %>%
  setNames(c("Misc Finish", "Count"))
kable(misc_table, row.names = F, align='l')
```


### Wins by Injury

<br>

```{r fig.width= 10, fig.height = 6, echo = FALSE, warning = FALSE, comment=FALSE}
injuries <- setDT(fights) %>% 
  na.omit() %>%
  dplyr::select(Category, method_d, round) %>%
  filter(Category == "Injury") %>%
  count(method_d, round, sort = T) %>%
  mutate(Round = round)

injuries_order <- aggregate(injuries$n, by=list(Injuries = injuries$method_d), FUN=sum) %>% 
  arrange(x) %>% pull(Injuries)

ggplot(injuries, aes(x=reorder(method_d,n), y=n, fill=Round)) +
  geom_bar(aes(fill=as.factor(Round)), stat='identity', position = position_stack(reverse = TRUE), col = 'black', width=0.85) + 
  coord_flip() + labs(x="Type of injury", y="Count") + ggtitle("Types of Injury Stoppages") +
  scale_fill_manual(values = sunshine_over_glacier[c(1,3,5)]) + theme_minimal() +
  scale_x_discrete(limits=injuries_order)


injuries_table <- aggregate(injuries$n, by=list(injuries = injuries$method_d), FUN=sum) %>% 
  arrange(desc(x)) %>%
  setNames(c("Injury", "Count"))
kable(injuries_table, row.names = F, align='l')
```
